FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT `EAGLE'

### 2015-12-17 source/33/TIGLIO.tex
no indent added

Nearly all Greek and Latin epigraphic texts are available, sometimes in multiple versions, after three decades of continuous digitization and online publication. Without counting repeated inscriptions and \textit{instrumenta}, there are about 300.000 Latin inscriptions\footnote{EAGLE disambiguated total: 235.626 ; EDCS not disambiguated total without \textit{instrumenta} and \textit{inscriptiones christianae}: 308.581. On the definition of Roman Epigraphy, see \citet{Panciera2012}.} online and about 200.000 Greek Inscriptions\footnote{Data from the latest Integrating Digital Epigraphies's (\url{https://youtu.be/OPfDj_hjeok}) harvest from the Packard Humanities Institute, Searchable Greek Inscriptions (\url{http://epigraphy.packhum.org/}), with some duplicates, 207.964.}. In some databases there is also abundant metadata and  a structured bibliography. The situation for translations and images of these inscriptions (text and support) available in the digital space is nevertheless quite different. The ratio between images of inscriptions and text is of one image every two inscriptions,\footnote{At the time in which this paper is being written (\today) the Epigraphic Database Heidelberg has ca. 35.000 photos and ca. 71.000 texts (0,5); The Epigraphic Database Roma has a slightly better ratio with 45.000 photos for 71.000 texts (0,6); The Epigraphic Database Bari has 34538 texts and 10341 images (0,3). In the EAGLE aggregator, the total ratio (excluding the related content of Arachne), is of 0,79 images for each text (235626 documental entities per 185999 visual entities) because smaller corpora tend to have a better photographic documentation.} but those inscriptions which have photographic documentation usually have many photos.\footnote{The Epigraphic Database Heidelberg has to date ca. 14.000 records with a photo or a drawing attached, bringing the average number of images per inscription to 2,5.} Translations are present only in smaller corpora edited online in most cases and the Attic Inscriptions Online project\footnote{\url{https://www.atticinscriptions.com}} is an \textit{uniquum} in its intent to provide mainly translations of inscriptions.\footnote{\citet{Lambert2014}} There are many publications offering translations in print, but these are not published online. An estimate calculation, based on the 11.000 translations present in the EAGLE Media Wiki, and known collections of printed translations of inscriptions, compared to a total of texts usefully translatable of around 300,000 texts, brings to an alarming 10\% of translated texts, of which only a third (slightly more than 3\%) is online. Translations are perhaps not a priority for researchers who know Greek and Latin, but are a way to clarify the interpretation of a text and an invaluable tool for didactical purposes and teaching: they are the only way in which an inscription can reach a wider public in a significant way as part of cultural heritage. The same could be said for images, even more obviously, since researchers also need them because: 1) they cannot always reach the place where an inscription is stored to study it  (given that the inscription is still there); 2) there are cases  in which a photo might be all we are left with and these are quickly increasing as monuments get lost or are destroyed. The imbalance in the documentation is thus pressing, since translations and images are our two best controls on the constitution and interpretation of ancient documentary texts. To an extent, digital epigraphy today is the direct descendant of epigraphy’s 19th century analog self: many texts, few translations, few images. This project aims to take initial steps to redress that imbalance, building resources that allow epigraphists and ancient historians to bring translations and images more closely into the suite of existing digital epigraphy resources.

@@

Let's look at the data we have. The EAGLE project has gathered some insights on small collections of images of inscriptions openly published online, on Wikimedia Commons,\footnote{\url{https://commons.wikimedia.org/wiki/Category:Media_contributed_by_EAGLE}} and on a set of translations of inscriptions, collected in the EAGLE Media Wiki.\footnote{\citet{Liuzzo2014} \url{http://www.eagle-network.eu/wiki/index.php/Main_Page}} We shall compare their impact and reach on the wider public to that of the texts of inscriptions, to underline the urgency for these materials to be produced also in order to bring epigraphy outside its restricted circles.

@@

\section{Translations}
To face these challenges with regard to translations, it is the opinion of the project team that tying together existing resources is a better way to tackle the issues rather than trying to superimpose a new tool or system. The available building blocks for such systematization of existing resources are the following:

@@

\section{Images}
Most of the existing images of inscriptions are currently safely stored in private computers. Large collections of images of inscriptions are available at the major databases, and can count up to thousands of photos of inscriptions or drawings. Large collections of images are also on Flickr (e.g. Visible Words) and Wikimedia Commons (CIL and AE categories provide a good overview of what is available).

@@

\section{User involvement and expert sourcing}
The problems of all digital projects looking at putting together materials from various sources and contributors are on one side to get people involved, on the other to overlook the work done and take care of the administration.  
The possible user scenarios are eventually many more. Two especially deserve to be mentioned here. Most input has been provided in term of new translations in the EAGLE Media Wiki during Workshops and Secondary Schools class work. Some teachers with a background in epigraphy have been contacted by the Epigraphic Database Rome to start and experiment with a didactic model which would include translating inscriptions. The students worked on a specific corpus of inscriptions, studied the text and the support and produced a translation which they entered in the EAGLE Media Wiki with the supervision of their teacher. This experience has proved successful for the students which have seen their contribution directly where it should be, together with other scholarly content. On the other side, every new translation matters in such a state as the one described above. 
The second example is the work done in two consecutive workshops, held in Ercolano and at the Centre for Hellenic Studies in Washington DC of the Ancient Graffiti Project,\footnote{\url{http://ancientgraffiti.wlu.edu/}} which has published multiple translations for all the graffiti of Herculaneum. In this context the usability of the Wikibase software has been tested and it proved to be an extremely intuitive and powerful tool. It takes very little explanation, but there are caveats for this simplicity and namely that it is extremely easy to do things in slightly creative ways, as entering statements as source information or typing an id in a slightly different way, which then need to be monitored and fixed.

\section{Conclusions}
Some of the tasks above have been already carried out, some are under way, but there are some general conclusions which can be summarized. There is a need to unify the resources, agree on standards for reference and citation and provide stable identifiers and citation structures, providing a comprehensive list of epigraphic publications with the relevant abbreviation in use. Although some work has been done, the amount of data makes this task continuously needed together with that of disambiguation. Other efforts need to go in the direction of flexible but harmonized standards for encoding and working on data entry giving priorities probably in a  slightly different way as before, updating the tools to be able to cope. More generally, while tools are abundant and so are guidelines and cookbooks, an agreed venue for coordination of the efforts is still a desideratum, and should include not only scholarly project but also community based efforts such as those of the Wikimedia Commons users and of the Flickr user's groups. These people could be also part of the peer review process, thus contrasting the side effects of an inactive board.\footnote{The last meeting of the project, dealing with images and CTS URN structure still had to take place at the date of submission of the present contribution.}

\section*{Aknowledgements}
The Andrew W. Mellon Foundation has funded the meetings of the project TIGLIO and the work which will be carried out for Attic Inscriptions Online. We would like to thank also all the Wikimedia users who have contributed to parts of this work in various ways and especially Aubrey, CristianCantoro, Fæ, Laurentius, Magnus Manske, Sannita and Wittylama. Another special aknowledgement goes to all users of the EAGLE Mediawiki and to all the volunteers working on photos of inscriptions online.

>>>>

\noindent Nearly all Greek and Latin epigraphic texts are available, sometimes in multiple versions, after three decades of continuous digitization and online publication. Without counting repeated inscriptions and \textit{instrumenta}, there are about 300.000 Latin inscriptions\footnote{EAGLE disambiguated total: 235.626 ; EDCS not disambiguated total without \textit{instrumenta} and \textit{inscriptiones christianae}: 308.581. On the definition of Roman Epigraphy, see \citet{Panciera2012}.} online and about 200.000 Greek Inscriptions\footnote{Data from the latest Integrating Digital Epigraphies's (\url{https://youtu.be/OPfDj_hjeok}) harvest from the Packard Humanities Institute, Searchable Greek Inscriptions (\url{http://epigraphy.packhum.org/}), with some duplicates, 207.964.}. In some databases there is also abundant metadata and  a structured bibliography. The situation for translations and images of these inscriptions (text and support) available in the digital space is nevertheless quite different. The ratio between images of inscriptions and text is of one image every two inscriptions,\footnote{At the time in which this paper is being written (\today) the Epigraphic Database Heidelberg has ca. 35.000 photos and ca. 71.000 texts (0,5); The Epigraphic Database Roma has a slightly better ratio with 45.000 photos for 71.000 texts (0,6); The Epigraphic Database Bari has 34538 texts and 10341 images (0,3). In the EAGLE aggregator, the total ratio (excluding the related content of Arachne), is of 0,79 images for each text (235626 documental entities per 185999 visual entities) because smaller corpora tend to have a better photographic documentation.} but those inscriptions which have photographic documentation usually have many photos.\footnote{The Epigraphic Database Heidelberg has to date ca. 14.000 records with a photo or a drawing attached, bringing the average number of images per inscription to 2,5.} Translations are present only in smaller corpora edited online in most cases and the Attic Inscriptions Online project\footnote{\url{https://www.atticinscriptions.com}} is an \textit{uniquum} in its intent to provide mainly translations of inscriptions.\footnote{\citet{Lambert2014}} There are many publications offering translations in print, but these are not published online. An estimate calculation, based on the 11.000 translations present in the EAGLE Media Wiki, and known collections of printed translations of inscriptions, compared to a total of texts usefully translatable of around 300,000 texts, brings to an alarming 10\% of translated texts, of which only a third (slightly more than 3\%) is online. Translations are perhaps not a priority for researchers who know Greek and Latin, but are a way to clarify the interpretation of a text and an invaluable tool for didactical purposes and teaching: they are the only way in which an inscription can reach a wider public in a significant way as part of cultural heritage. The same could be said for images, even more obviously, since researchers also need them because: 1) they cannot always reach the place where an inscription is stored to study it  (given that the inscription is still there); 2) there are cases  in which a photo might be all we are left with and these are quickly increasing as monuments get lost or are destroyed. The imbalance in the documentation is thus pressing, since translations and images are our two best controls on the constitution and interpretation of ancient documentary texts. To an extent, digital epigraphy today is the direct descendant of epigraphy’s 19th century analog self: many texts, few translations, few images. This project aims to take initial steps to redress that imbalance, building resources that allow epigraphists and ancient historians to bring translations and images more closely into the suite of existing digital epigraphy resources.

@@

\noindent Let's look at the data we have. The EAGLE project has gathered some insights on small collections of images of inscriptions openly published online, on Wikimedia Commons,\footnote{\url{https://commons.wikimedia.org/wiki/Category:Media_contributed_by_EAGLE}} and on a set of translations of inscriptions, collected in the EAGLE Media Wiki.\footnote{\citet{Liuzzo2014} \url{http://www.eagle-network.eu/wiki/index.php/Main_Page}} We shall compare their impact and reach on the wider public to that of the texts of inscriptions, to underline the urgency for these materials to be produced also in order to bring epigraphy outside its restricted circles.

@@

\section{Translations}
\noindent To face these challenges with regard to translations, it is the opinion of the project team that tying together existing resources is a better way to tackle the issues rather than trying to superimpose a new tool or system. The available building blocks for such systematization of existing resources are the following:

@@

\section{Images}
\noindent Most of the existing images of inscriptions are currently safely stored in private computers. Large collections of images of inscriptions are available at the major databases, and can count up to thousands of photos of inscriptions or drawings. Large collections of images are also on Flickr (e.g. Visible Words) and Wikimedia Commons (CIL and AE categories provide a good overview of what is available).

@@

\section{User involvement and expert sourcing}
\noindent The problems of all digital projects looking at putting together materials from various sources and contributors are on one side to get people involved, on the other to overlook the work done and take care of the administration.  
The possible user scenarios are eventually many more. Two especially deserve to be mentioned here. Most input has been provided in term of new translations in the EAGLE Media Wiki during Workshops and Secondary Schools class work. Some teachers with a background in epigraphy have been contacted by the Epigraphic Database Rome to start and experiment with a didactic model which would include translating inscriptions. The students worked on a specific corpus of inscriptions, studied the text and the support and produced a translation which they entered in the EAGLE Media Wiki with the supervision of their teacher. This experience has proved successful for the students which have seen their contribution directly where it should be, together with other scholarly content. On the other side, every new translation matters in such a state as the one described above. 
The second example is the work done in two consecutive workshops, held in Ercolano and at the Centre for Hellenic Studies in Washington DC of the Ancient Graffiti Project,\footnote{\url{http://ancientgraffiti.wlu.edu/}} which has published multiple translations for all the graffiti of Herculaneum. In this context the usability of the Wikibase software has been tested and it proved to be an extremely intuitive and powerful tool. It takes very little explanation, but there are caveats for this simplicity and namely that it is extremely easy to do things in slightly creative ways, as entering statements as source information or typing an id in a slightly different way, which then need to be monitored and fixed.

\section{Conclusions}
\noindent Some of the tasks above have been already carried out, some are under way, but there are some general conclusions which can be summarized. There is a need to unify the resources, agree on standards for reference and citation and provide stable identifiers and citation structures, providing a comprehensive list of epigraphic publications with the relevant abbreviation in use. Although some work has been done, the amount of data makes this task continuously needed together with that of disambiguation. Other efforts need to go in the direction of flexible but harmonized standards for encoding and working on data entry giving priorities probably in a  slightly different way as before, updating the tools to be able to cope. More generally, while tools are abundant and so are guidelines and cookbooks, an agreed venue for coordination of the efforts is still a desideratum, and should include not only scholarly project but also community based efforts such as those of the Wikimedia Commons users and of the Flickr user's groups. These people could be also part of the peer review process, thus contrasting the side effects of an inactive board.\footnote{The last meeting of the project, dealing with images and CTS URN structure still had to take place at the date of submission of the present contribution.}

\section*{Aknowledgements}
\noindent The Andrew W. Mellon Foundation has funded the meetings of the project TIGLIO and the work which will be carried out for Attic Inscriptions Online. We would like to thank also all the Wikimedia users who have contributed to parts of this work in various ways and especially Aubrey, CristianCantoro, Fæ, Laurentius, Magnus Manske, Sannita and Wittylama. Another special aknowledgement goes to all users of the EAGLE Mediawiki and to all the volunteers working on photos of inscriptions online.

### 2015-12-17 source/36/DatabasesToEpiDoc.tex
no indent added

The Europeana network of Ancient Greek and Latin Epigraphy\footnote{\citet{OrlandiCasarosa2014} \citet{Orlandi2014} \citet{liuzzo2014} \url{http://www.eagle-network.eu/}} brings
together most repositories of ancient epigraphic material and aims to
provide historians not just with a useful research tool, but a
curated online edition which has high quality contents as well as high
quality data. Towards this end many steps are needed as the databases on which many institutions have worked for decades cannot be just discarded. Epigraphy enjoys a special position among digital resources as most transcriptions of existing classical inscriptions have been digitized. Many duplicates exist and a lot of the content lacks metadata almost entirely, but the work continues with dedication so that among the 10 \% of digitised European Cultural Heritage, Ancient Epigraphy can certainly claim to play a strong role.\footnote{The data is taken from the Europeana 2020 strategy report available at \url{http://strategy2020.europeana.eu/}.} While the long term aim for epigraphy is on the one hand to keep up with new finds and studies and on the other to have a common and flexible entry point and backend on which to work together, this is not easy to accomplish without a careful process which involves a gradual transition from independent databases (the best option at the time when the EAGLE databases were created) to a common online data entry and editing system working on XML files in EpiDoc, as it has now become feasible.   
In this paper some of the mapping and content harmonization efforts undertaken by the EAGLE BPN to achieve this step towards a common epigraphic resource, will be presented as a case in which this apparently technical task involved deep revision of contents related to the discipline and required discussions and collaborative efforts by the working
groups of digital epigraphists, encoders, epigraphists and developers, coming from several different background and experiences. The result of this work are the automated conversion to TEI EpiDoc XML of 90\% of the text features in the EAGLE databases and the EAGLE Vocabularies, the largest existing controlled and aligned vocabularies for classical epigraphy.\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}}

\section{Why XML}
Choosing XML for an online publication is almost obvious given the current stage of development of the editorial methodology in epigraphy. However there is room for a brief explanation of why this is preferable on a large scale to a relational database. I would like to suggest a comparison here. Coffee is produced by filtering water through a powder more or less thick. This can be accomplished by pushing water through the coffee (as in most systems, from a mocha to an espresso machine) or coffee through the water (as in a french press, where the coffee floats and is pushed down). There is pretty much the same difference when we compare the action of entering data into a database, i.e. a rigid structure of information, and marking up a text, adding tags to the text itself. The first process forces data through a grid however well thought off, while in the second case structure is added on the contents without altering them, but just enhancing their potential so that the richness and values of both contents and structure empower each other. There is just more taste and much more freedom of encoding to the level desired and to the complexity one wants. An additional advantage is the level of interoperability. An XML file is virtually software independent, it is not an Access, FileMaker or whatever database format. A third reason is that an XML file can host a complete textual description with semantic markup and still have the structure of a normal text edition, while the same source can be used to produce a database or an index for interrogation purposes. For example the same markup of an abbreviation \texttt{<expan><abbr>C</abbr><ex>aio</ex></expan>} can be used to produce a database of abbreviation marks and expansions, to output in a diplomatic edition only the character actually on the stone or to print the expanded abbreviation in a critical edition.\footnote{\url{http://www.stoa.org/epidoc/gl/latest/trans-abbrevmark.html}} To print a database to a book would be at least more cumbersome. XML provides also a better service as an archival format being explicit on the values it uses to describe contents, moreover if these follow a schema of agreed value of such elements. This said producing and sharing XML files means that more quality editions can be produced, printed editions can be produced and all sort of outputs can be supported with flexibility; it means that data is safe from software developments and support; it means you don't have to put information into a rigid table which is then hard to restructure, but that you can annotate your text as much as you want thus adding layers and layers of ``possible databases'' of relevant information to be extracted and used.   
The text is what one uses to describe situations and complexities, the markup allows also the machine to know and retrieve that information as in a database: you don't need a field or an element to state what the orientation of your text is, you can simply state and describe as you would do in a normal publication and if you want to reuse information about the text orientation then you can add elements and attributes to do so. XML allows not to worry too much on ``what goes where'' or about ``where do I put that info" because it is not a database, so that the editor can instead focus on ``what is that'' and on accurately describing it. The TEI specification which is EpiDoc allows the specific epigraphic content to be described fully for what concerns the text of inscriptions, drawing on more than a decade long experimenting and development of this standard thanks to a vast and active international community. The simple fact of using the same markup to describe our texts allows us to study them together and to produce and run software for more then just one corpus of inscriptions.

\section{EAGLE Mappings and Metadata modeling}
The EAGLE BPN has taken 18 different content providers and has come to
the decision to take the first steps to bring hundreds of thousands of
ancient Greek and Roman inscriptions to TEI. This required 14 mappings
of local database metadata models to TEI/EpiDoc, as well as the elaboration of
XSL Transformations\footnote{\url{https://github.com/EAGLE-BPN}} for up-conversion\footnote{This is the terminology used by \citet[906]{Kay} to describe transforming without explicit structure in data which has it. In this case it is a transformation from a less explicit database structure to a fully explicit and interoperable one.} and the alignment of the text
markup. Although this was thought to be a trivial task it turned out
that whereas databases claimed to use the same conventions, a considerable amount of differentiation took place due to the data structure and to data entry procedures as well as to policies and internal decisions.

@@

The occasion of a mapping work allows also for other tasks of curation to be performed.
EAGLE Members had in some case the text of a same inscription with three parallel different types of encoding conventions applied.

@@

\subsection{Step 1}
Each project uses different conventions and therefore the regular expressions used to match particular situations are different.
The process of mark up of the string text in div[@type="edition"] is accomplished in several steps to guarantee consistency and precision.

@@

\subsection{Step 2}
On the elements which contain information such as Object Type, Material, Execution Technique, etc. which are typically handled with a controlled vocabulary a series of XSL transformations is passed, one for each vocabulary related to that element  to match the content of the element with the vocabulary entry into the SKOS vocabulary stored in git, regularly updated and published as a self standing resource on the EAGLE portal.\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}} What follows is a brief description of such vocabularies and their development.

@@

\section{Linked Open Data}
The Europeana LOD practice for metadata recommends the adoption of machine readable vocabularies. Within the EAGLE BPN Linked Open Data practices and approaches have been taken on board for a process which will bring current material some steps closer to that practice.\footnote{\citet{Bizer2009}} Addressing existing problems in classification and the publication of a machine readable vocabulary of values (controlled vocabulary) is one of these steps:\footnote{\citet[4-5]{Harper2012} for the definition. See also \citet{Isaac2012}} it involves identifying and matching values as well as giving them stable identifiers and making of them self standing resources where possible.

@@

\section{Crosswalking EpiDoc to EDH with XSLT}
The proof of concept of usability of the XML source file is given by the reverse function being possible, i.e. to transform the XML files into a database data structure. A team of people from the EAGLE BPN\footnote{Gabriel Bodard, Pietro Liuzzo, James Cowey, Frank Grieshaber, Brigitte Gräf and Francisca Feraudi Gruénais.} and Scott Vanderbilt, curator of the online edition of the Roman Inscriptions of Britain,\footnote{\url{http://romaninscriptionsofbritain.org/}} worked on mapping the EpiDoc contents of that project to the EDH database model,\footnote{The work was largely based on the previous exercise of this kind, made by Gabriel Bodard and James Cowey for the Inscriptions of Roman Tripolitania.} facing a series of problems and challenges, which brought an interesting discussion, also on principles of ``up-conversion'' and cross-walking, which should be as often as possible multi-directional.\footnote{The files are available here: \url{https://raw.githubusercontent.com/EAGLE-BPN/Epidoc2EDH/master/rib_to_edh-2.xsl}}

@@

\section{Conclusion}
I have presented in this paper the effort of a large consortium of people and institutions belonging to the same field of interest, that of Ancient Roman and Greek Epigraphy.
Mappings and harmonization of data have been for the EAGLE project not just a way to allow machines to do more, but an intellectual effort of revision of contents and establishment of best practices. This work prepares further steps which hopefully will take place soon, in which one repository of inscriptions will be accessible to all to edit, contribute and download data from several websites and interfaces. In such environments related contents based on descriptive URIs, as well as places, personal names and other semantic information will be possible and meaningful and perhaps in a not too far future classicists will have a complete and functioning network of Linked Ancient World Data, including texts, manuscripts, papyri, coins and inscriptions.

>>>>

\noindent The Europeana network of Ancient Greek and Latin Epigraphy\footnote{\citet{OrlandiCasarosa2014} \citet{Orlandi2014} \citet{liuzzo2014} \url{http://www.eagle-network.eu/}} brings
together most repositories of ancient epigraphic material and aims to
provide historians not just with a useful research tool, but a
curated online edition which has high quality contents as well as high
quality data. Towards this end many steps are needed as the databases on which many institutions have worked for decades cannot be just discarded. Epigraphy enjoys a special position among digital resources as most transcriptions of existing classical inscriptions have been digitized. Many duplicates exist and a lot of the content lacks metadata almost entirely, but the work continues with dedication so that among the 10 \% of digitised European Cultural Heritage, Ancient Epigraphy can certainly claim to play a strong role.\footnote{The data is taken from the Europeana 2020 strategy report available at \url{http://strategy2020.europeana.eu/}.} While the long term aim for epigraphy is on the one hand to keep up with new finds and studies and on the other to have a common and flexible entry point and backend on which to work together, this is not easy to accomplish without a careful process which involves a gradual transition from independent databases (the best option at the time when the EAGLE databases were created) to a common online data entry and editing system working on XML files in EpiDoc, as it has now become feasible.   
In this paper some of the mapping and content harmonization efforts undertaken by the EAGLE BPN to achieve this step towards a common epigraphic resource, will be presented as a case in which this apparently technical task involved deep revision of contents related to the discipline and required discussions and collaborative efforts by the working
groups of digital epigraphists, encoders, epigraphists and developers, coming from several different background and experiences. The result of this work are the automated conversion to TEI EpiDoc XML of 90\% of the text features in the EAGLE databases and the EAGLE Vocabularies, the largest existing controlled and aligned vocabularies for classical epigraphy.\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}}

\section{Why XML}
\noindent Choosing XML for an online publication is almost obvious given the current stage of development of the editorial methodology in epigraphy. However there is room for a brief explanation of why this is preferable on a large scale to a relational database. I would like to suggest a comparison here. Coffee is produced by filtering water through a powder more or less thick. This can be accomplished by pushing water through the coffee (as in most systems, from a mocha to an espresso machine) or coffee through the water (as in a french press, where the coffee floats and is pushed down). There is pretty much the same difference when we compare the action of entering data into a database, i.e. a rigid structure of information, and marking up a text, adding tags to the text itself. The first process forces data through a grid however well thought off, while in the second case structure is added on the contents without altering them, but just enhancing their potential so that the richness and values of both contents and structure empower each other. There is just more taste and much more freedom of encoding to the level desired and to the complexity one wants. An additional advantage is the level of interoperability. An XML file is virtually software independent, it is not an Access, FileMaker or whatever database format. A third reason is that an XML file can host a complete textual description with semantic markup and still have the structure of a normal text edition, while the same source can be used to produce a database or an index for interrogation purposes. For example the same markup of an abbreviation \texttt{<expan><abbr>C</abbr><ex>aio</ex></expan>} can be used to produce a database of abbreviation marks and expansions, to output in a diplomatic edition only the character actually on the stone or to print the expanded abbreviation in a critical edition.\footnote{\url{http://www.stoa.org/epidoc/gl/latest/trans-abbrevmark.html}} To print a database to a book would be at least more cumbersome. XML provides also a better service as an archival format being explicit on the values it uses to describe contents, moreover if these follow a schema of agreed value of such elements. This said producing and sharing XML files means that more quality editions can be produced, printed editions can be produced and all sort of outputs can be supported with flexibility; it means that data is safe from software developments and support; it means you don't have to put information into a rigid table which is then hard to restructure, but that you can annotate your text as much as you want thus adding layers and layers of ``possible databases'' of relevant information to be extracted and used.   
The text is what one uses to describe situations and complexities, the markup allows also the machine to know and retrieve that information as in a database: you don't need a field or an element to state what the orientation of your text is, you can simply state and describe as you would do in a normal publication and if you want to reuse information about the text orientation then you can add elements and attributes to do so. XML allows not to worry too much on ``what goes where'' or about ``where do I put that info" because it is not a database, so that the editor can instead focus on ``what is that'' and on accurately describing it. The TEI specification which is EpiDoc allows the specific epigraphic content to be described fully for what concerns the text of inscriptions, drawing on more than a decade long experimenting and development of this standard thanks to a vast and active international community. The simple fact of using the same markup to describe our texts allows us to study them together and to produce and run software for more then just one corpus of inscriptions.

\section{EAGLE Mappings and Metadata modeling}
\noindent The EAGLE BPN has taken 18 different content providers and has come to
the decision to take the first steps to bring hundreds of thousands of
ancient Greek and Roman inscriptions to TEI. This required 14 mappings
of local database metadata models to TEI/EpiDoc, as well as the elaboration of
XSL Transformations\footnote{\url{https://github.com/EAGLE-BPN}} for up-conversion\footnote{This is the terminology used by \citet[906]{Kay} to describe transforming without explicit structure in data which has it. In this case it is a transformation from a less explicit database structure to a fully explicit and interoperable one.} and the alignment of the text
markup. Although this was thought to be a trivial task it turned out
that whereas databases claimed to use the same conventions, a considerable amount of differentiation took place due to the data structure and to data entry procedures as well as to policies and internal decisions.

@@

\noindent The occasion of a mapping work allows also for other tasks of curation to be performed.
EAGLE Members had in some case the text of a same inscription with three parallel different types of encoding conventions applied.

@@

\subsection{Step 1}
\noindent Each project uses different conventions and therefore the regular expressions used to match particular situations are different.
The process of mark up of the string text in div[@type="edition"] is accomplished in several steps to guarantee consistency and precision.

@@

\subsection{Step 2}
\noindent On the elements which contain information such as Object Type, Material, Execution Technique, etc. which are typically handled with a controlled vocabulary a series of XSL transformations is passed, one for each vocabulary related to that element  to match the content of the element with the vocabulary entry into the SKOS vocabulary stored in git, regularly updated and published as a self standing resource on the EAGLE portal.\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}} What follows is a brief description of such vocabularies and their development.

@@

\section{Linked Open Data}
\noindent The Europeana LOD practice for metadata recommends the adoption of machine readable vocabularies. Within the EAGLE BPN Linked Open Data practices and approaches have been taken on board for a process which will bring current material some steps closer to that practice.\footnote{\citet{Bizer2009}} Addressing existing problems in classification and the publication of a machine readable vocabulary of values (controlled vocabulary) is one of these steps:\footnote{\citet[4-5]{Harper2012} for the definition. See also \citet{Isaac2012}} it involves identifying and matching values as well as giving them stable identifiers and making of them self standing resources where possible.

@@

\section{Crosswalking EpiDoc to EDH with XSLT}
\noindent The proof of concept of usability of the XML source file is given by the reverse function being possible, i.e. to transform the XML files into a database data structure. A team of people from the EAGLE BPN\footnote{Gabriel Bodard, Pietro Liuzzo, James Cowey, Frank Grieshaber, Brigitte Gräf and Francisca Feraudi Gruénais.} and Scott Vanderbilt, curator of the online edition of the Roman Inscriptions of Britain,\footnote{\url{http://romaninscriptionsofbritain.org/}} worked on mapping the EpiDoc contents of that project to the EDH database model,\footnote{The work was largely based on the previous exercise of this kind, made by Gabriel Bodard and James Cowey for the Inscriptions of Roman Tripolitania.} facing a series of problems and challenges, which brought an interesting discussion, also on principles of ``up-conversion'' and cross-walking, which should be as often as possible multi-directional.\footnote{The files are available here: \url{https://raw.githubusercontent.com/EAGLE-BPN/Epidoc2EDH/master/rib_to_edh-2.xsl}}

@@

\section{Conclusion}
\noindent I have presented in this paper the effort of a large consortium of people and institutions belonging to the same field of interest, that of Ancient Roman and Greek Epigraphy.
Mappings and harmonization of data have been for the EAGLE project not just a way to allow machines to do more, but an intellectual effort of revision of contents and establishment of best practices. This work prepares further steps which hopefully will take place soon, in which one repository of inscriptions will be accessible to all to edit, contribute and download data from several websites and interfaces. In such environments related contents based on descriptive URIs, as well as places, personal names and other semantic information will be possible and meaningful and perhaps in a not too far future classicists will have a complete and functioning network of Linked Ancient World Data, including texts, manuscripts, papyri, coins and inscriptions.

### 2015-12-17 source/38/EAGLE2016_Final_Paper Modolo.tex
no indent added

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Mirco Modolo}
\date{2015-11-20}
\begin{document}
Abstract

\begin{figure}
\centering
\begin{minipage}{10.901cm}
FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT
‘EAGLE’\newline

Mirco Modolo

Free lance archaeologist and archivist

Email(s): mircomodolo@gmail.com
\end{minipage}
\end{figure}
This paper traces the genesis and activity of the “Fotografie libere per i Beni Culturali” movement promoting free
remote reproduction, and reuse for scientific purposes, of manuscripts and antique volumes conserved in Italian
archives and libraries. Simultaneously, evidence highlights the advantages of deregulating cultural heritage images for
historical research, in particular for the study of antique epigraphy developed by the EAGLE project. \

\bigskip

Keywords

Archivi, Biblioteche, Manoscritti, Riproduzioni, Creative Commons

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\begin{enumerate}
\item Free reproduction of cultural heritage outlined in the “Art Bonus” decree
\end{enumerate}
\ L’entrata in vigore, il primo giugno 2014, del decreto “Art Bonus”, che sanciva la libera riproduzione per finalità
scientifiche, purché effettuata a distanza, di tutti i beni culturali rappresentò una novità assoluta per la ricerca
storica: numerosi archivi e biblioteche resero infatti libera e gratuita la riproduzione con la fotocamera dell’utente
di manoscritti e volumi storici, con evidente vantaggio per la ricerca storica e in particolare per lo studio
dell’epigrafia nei manoscritti antiquari.

@@

\bigskip

@@

\bigskip

\bigskip

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT `EAGLE'}
\author[free]{Mirco Modolo\corref{first}}
\address[free]{Free lance archaeologist and archivist}
\cortext[first]{Corresponding author. Email: mircomodolo@gmail.com}
\date{2015-11-20}
\begin{document}

\maketitle
\begin{abstract}
This paper traces the genesis and activity of the “Fotografie libere per i Beni Culturali” movement promoting free
remote reproduction, and reuse for scientific purposes, of manuscripts and antique volumes conserved in Italian
archives and libraries. Simultaneously, evidence highlights the advantages of deregulating cultural heritage images for
historical research, in particular for the study of antique epigraphy developed by the EAGLE project.
\end{abstract}
\keywords{Archivi, Biblioteche, Manoscritti, Riproduzioni, Creative Commons}

Free reproduction of cultural heritage outlined in the “Art Bonus” decree

L’entrata in vigore, il primo giugno 2014, del decreto “Art Bonus”, che sanciva la libera riproduzione per finalità
scientifiche, purché effettuata a distanza, di tutti i beni culturali rappresentò una novità assoluta per la ricerca
storica: numerosi archivi e biblioteche resero infatti libera e gratuita la riproduzione con la fotocamera dell’utente
di manoscritti e volumi storici, con evidente vantaggio per la ricerca storica e in particolare per lo studio
dell’epigrafia nei manoscritti antiquari.

@@

@@
